---
title: 'Tornado outbreak climatology in the United States (1995 - 2019): Definitions,
  description, and statistical analyses'
author: "Zoe Schroder"
date: "7/20/2020"
output: html_document
editor_options: 
  chunk_output_type: inline
---

Load the libraries needed for this project. 
```{r}
suppressMessages(library(sf))
suppressMessages(library(dplyr))
#install.packages("lubridate")
suppressMessages(library(lubridate))
suppressMessages(library(USAboundaries))
#devtools::install_github("ropensci/USAboundariesData")
suppressMessages(library(USAboundariesData))
suppressMessages(library(tmap))
suppressMessages(library(ggplot2))
```

## Tornado-Level Statistics: 

The newest GIS shapefile contains missing geometries for more than 30% of the tornadoes. The number of missing geometries is highest after 1995. Instead here we use the csv file from https://www.spc.noaa.gov/wcm/#data  Use the start lon/lat and create a `sp` object then convert to `sf`. Set the coordinate reference system (crs) to ESPG 4326.
```{r}
Torn.spdf <- read.csv(file = "1950-2018_actual_tornadoes.csv")
sp::coordinates(Torn.spdf) <- ~ slon + slat
Torn.sfdf <- st_as_sf(Torn.spdf)
st_crs(Torn.sfdf) <- 4326
```

Remove tornadoes in Hawaii, Alaska, and Puerto Rico and those occurring before 1994. That year marks the beginning of comprehensive WSR-88D radar. For missing EF ratings use the modification rules (if/else) defined here: https://www.spc.noaa.gov/wcm/OneTor_F-scale-modifications.pdf 
```{r}
All_Tornadoes <- Torn.sfdf %>%
  filter(yr >= 1994,
         !st %in% c("AK", "PR", "HI")) %>%
  mutate(mag = ifelse(mag == -9 & len <= 5, 0, mag),
         mag = ifelse(mag == -9 & len > 5, 1, mag))
```

Add a data/time column also add columns for path length, width, and area in metric units. Leave the time zone as native CDT. Create a convective day (6AM to 6AM) column taking hours 00:00:00 -> 05:59:59 and assigning it to the previous date (this associates the previous day's date to tornadoes occurring up to 6 hours after local midnight).
```{r}
All_Tornadoes <- All_Tornadoes %>%
  mutate(#dy = format(as.Date(date,format="%y/%m/%d"), "%d"), #This is now an included column in data
         DateTime = as.POSIXct(paste(yr, mo, dy, time), format = "%Y%m%d%H:%M:%S"),
         Hour = hour(DateTime),
         Year = year(DateTime),
         cDateTime = DateTime - as.difftime(6, unit = "hours"),
         cDate = as.Date(as_datetime(ifelse(Hour < 6, (DateTime - 86400), cDateTime), tz = Sys.timezone())),
         Length = len * 1609.34,
         Length = ifelse(Length == 0, min(Length[Length > 0]), Length), #takes care of zero length
         Width = wid * .9144,
         Width = ifelse(Width == 0, min(Width[Width > 0]), Width), #takes care of zero width
         Width = ifelse(Year >= 1995, Width * pi/4, Width), #takes care of change: avg to max
         cas = inj + fat,
         AreaPath = Length * Width,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(All_Tornadoes)[1]
```

The geometry type is `POINT`. Each tornado is represented as a single point location geometry (start location). Add power dissipation per tornado.

Add power dissipation per tornado. Use the empirical model for tornado winds by EF rating taken from Table 3-1 of NRC 2007. Percent area by EF rating for each EF category. Threshold wind speeds (m/s) are a lower bound 3-sec gusts on the operational EF Scale (Table 2-1 of NRC2007). This is based on work by Fricker et al. (2017). The model is
$$
E = A_p \rho \sum_{j=0}^{J} w_j v_j^{3},
$$
where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
ef <- All_Tornadoes$mag + 1
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
All_Tornadoes <- All_Tornadoes %>%
  mutate(ED = EW3 * AreaPath)
``` 

Determine the distance between tornadoes in space and time. Use a projection, not lat/lon. See https://epsg.io/102004. Extract the coordinates of the start locations as a N by 2 matrix, where N is the number of tornadoes. Also extract the date-time as a vector of class `POSIXct`.
```{r}
# Lambert Conic Conformal
All_Tornadoes <- st_transform(All_Tornadoes, crs = "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m no_defs")
space <- st_coordinates(All_Tornadoes)
time <- All_Tornadoes$DateTime
```

Next compute pairwise Euclidean distances in space and, separately, in time using the `dist()` function. Divide the spatial distance by 15 so that the values are commensurate with the time 'distance' based on the assumption of 15 meters per second (~34 mph) for an average speed of tornado-generating storms. Compare: Distance from New York to Denver is 2.622 x 10^6 meters. There are 3.154 x 10^7 seconds in a year. This will capture the historic multiday tornado outbreaks. For analysis we want to consider each day in the multiday group separately. As the value of the divisor increases cluster areas get larger. Remove `ds` and `dt` to free memory. Distances are saved as an object of class `dist` containing a vector of length N * (N-1)/2, which is the number of unique point pairs.
```{r}
ds <- dist(space) / 15
dt <- dist(time)
dst <- ds + dt
rm(ds, dt)
```

Distances are saved as an object of class `dist` containing a vector of length N * (N-1)/2, which is the number of unique point pairs.

Next group the tornadoes based on the space-time distances. This is done with the `hclust()` (hierarchical cluster) function. Initially, each tornado is assigned to its own group and then the algorithm joins the two closest tornadoes determined by values in `dst`. The algorithm continues by joining tornadoes (and tornado groups) until there is a single large group.

The single linkage method (`method = "single"`) is related to the minimal spanning tree (MST) and adopts a 'friends of friends' grouping strategy. An edge-weighted graph is a graph where each edge has a weight (or cost). Here weights are space-time distances between tornadoes. A MST of an edge-weighted graph is a spanning tree whose weight (the sum of the weights of its edges) is no larger than the weight of any other spanning tree. A spanning tree of a graph on N vertices (tornado centroids) is a subset of N-1 edges that form a tree (Skiena 1990, p. 227).
 
The `cutree()` function is used to extract a group number for each tornado. Tornadoes in each group are close in space & time. Here the tree is cut at a height of 50000 space-time units. Making `h` smaller results in smaller groups (fewer tornadoes per group).
```{r}
stime <- proc.time()
tree <- hclust(dst, method = "single")
groupNumber <- as.integer(cutree(tree, h = 50000))
proc.time() - stime
```

Add the group number to each tornado. 
```{r}
All_Tornadoes$groupNumber <- groupNumber
```

# Tornado Counts


`Generate the state and county borders`. Use the laea projection. 
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")]
stateBorders <- us_states(states = sts)
counties.sf <- us_counties()

tm_shape(Torn.sfdf) +
  tm_dots() + 
tm_shape(stateBorders, projection = "+proj=laea +lat_0=0 +lon_0=-90", is.master = TRUE) + 
  tm_borders()
#upside down triangles and colored by EF rating
```

State tornado counts from 1994 to 2018. 

```{r}
states.sf <- st_transform(stateBorders,  
                            crs = st_crs(All_Tornadoes))
gI <- st_intersects(All_Tornadoes, states.sf, sparse = FALSE)
colnames(gI) <- states.sf$name
states.sf$GroupDayCount <- colSums(gI)
```

Remove Alaska, Hawaii, and Puerto Rico from the `states.sf` data frame.
```{r}
states.sf <- states.sf %>% 
  filter(!stusps %in% c("AK", "PR", "HI"))
```

Create a map of the outbreak density in each county in the United States. Fill by the `GroupDayCount` column. 
```{r}
states.sf <- st_transform(states.sf, 
                            crs = st_crs(stateBorders))
All_Tornadoes <- st_transform(All_Tornadoes, 
                            crs = st_crs(stateBorders))
```

```{r}
Torn.sf <- as.data.frame(Torn.spdf)

Torn.sf <- Torn.sf %>%
  filter(yr >= 1994, 
         !st %in% c("AK", "PR", "HI"))%>%
  mutate(cas = fat + inj)

ggplot(Torn.sf, aes(x = slon, y = slat)) +
  geom_hex(bins = 50) +
  scale_fill_continuous(type = "viridis") +
  borders("state", region = c(stateBorders$state_name), col = "gray70") +
  theme_bw() +
  xlab(expression(paste('Longitude (',~degree,'W)',sep=''))) +
  ylab(expression(paste('Latitude (',~degree,'N)',sep='')))
```

# Monthly and Yearly Variation

*Year*
```{r}
TornsbyYear <- All_Tornadoes %>% 
  group_by(year = yr) %>%
  summarize(totaltornadoes = n())
```

`Get the week labels for the data`
```{r}
x.Date <- as.Date(paste(rep(1994:2018, each = 12), rep(1:12, 2), 1, sep = "-"))
library(zoo)
x <- zoo(rnorm(24), x.Date)
times <- time(x)
ticks <- as.data.frame(x = seq(times[1], times[length(times)], by = "weeks"))

week <- as.data.frame(ticks[1:53,])
  
months <- as.data.frame(format(week, "%b"))
Mo <- as.data.frame(format(week, "%m"))
day <- as.data.frame(format(week, "%d"))

dat <- as.data.frame(cbind(week, Mo, months, day))
colnames(dat) <- c("Week", "Mo", "Month", "Day") 

dat <- dat %>%
  group_by(Month, Day, Mo) %>%
  summarize(count = n(),
            Week = paste0(Month, " ", Day))

dat <- dat[order(dat$Mo),]
```

Make a figure of the change in tornadoes per annum. 

```{r}
yeargraph <- ggplot(TornsbyYear, (aes(x = year, y = totaltornadoes))) +
  geom_smooth(method = lm, span = .5, se = FALSE, color = "gray70", size = 1) +
  geom_line(color = "black", lwd = 1) +
  scale_x_continuous(expand = c(0, 0), breaks = c(seq(1994,2018,1))) +
  scale_y_continuous(limits = c(800, 2000), breaks = c(seq(800,2000,150))) + 
#  geom_smooth(aes(x = week, y = probs/100), span = .5, se = FALSE, color = "gray70", size = 1) +
  theme_bw() +
  xlab("Year") +
  ylab("Number of Tornadoes\n ")

yeargraph <- yeargraph + theme(axis.title.y = element_text(colour = "black"), axis.text.y = element_text(color = "black"),  axis.text.x = element_text(angle = 45, hjust = 1), axis.text = element_text(size = 12), axis.title=element_text(size=14))

yeargraph
```

*Month*
```{r}
TornsByMonth <- All_Tornadoes %>% 
  group_by(mo) %>%
  summarize(totaltornadoes = n()) 
```

`Make a figure of the total tornadoes from 1994 to 2018 by month`
```{r}
ggplot(TornsByMonth, aes(x = as.factor(mo), y = totaltornadoes)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_discrete(labels = month.name) +
  scale_y_continuous(limits = c(0, NA)) +
  ylab("Number of Tornado Groups\nWith At Least 10 Tornadoes") + xlab("") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

( p0 <- ggplot(TornsByMonth, aes(x = mo, y = totaltornadoes)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_continuous(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, 7000)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Tornadoes") +
  theme_minimal() )
```

## Daily (hourly) variation in tornado activity

```{r}
TornsbyHour <- All_Tornadoes %>% 
  group_by(TornHour = Hour) %>%
  summarize(totaltornadoes = n())
```

`Make the labels for the x-axis`
```{r}
hourlabs = c("12 AM", "1 AM", "2 AM", "3 AM", "4 AM", "5 AM", "6 AM", "7 AM", "8 AM", "9 AM", "10 AM", "11 AM", "12 PM", "1 PM","2 PM", "3 PM", "4 PM", "5 PM", "6 PM", "7 PM", "8 PM", "9 PM", "10 PM", "11 PM")

hourgraph <- ggplot(TornsbyHour, (aes(x = TornHour, y = totaltornadoes))) +
  #geom_smooth(aes(x = Tornhour, y = totaltornadoes), span = .5, se = FALSE, color = "gray70", size = 1) +
  geom_bar(stat = "identity", color = "gray70", fill = "gray70") +
  scale_x_continuous(expand = c(0, 0), breaks = c(seq(0,23,1)), labels = hourlabs) +
  scale_y_continuous(limits = c(0, 4000), breaks = c(seq(0,4000,500))) + 
#  geom_smooth(aes(x = week, y = probs/100), span = .5, se = FALSE, color = "gray70", size = 1) +
  theme_bw() +
  xlab("Hour") +
  ylab("Number of Tornadoes\n ")

hourgraph <- hourgraph + theme(axis.title.y = element_text(colour = "black"), axis.text.y = element_text(color = "black"),  axis.text.x = element_text(angle = 45, hjust = 1), axis.text = element_text(size = 12), axis.title=element_text(size=14))

hourgraph
```

`Make a circular bar graph`
```{r}
ggplot(TornsbyHour, (aes(x = TornHour, y = totaltornadoes))) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_continuous(breaks = seq(0, 23, 1), labels = hourlabs) +
  scale_y_continuous(limits = c(0, 4000)) +
  coord_polar(start = 0) +
  labs(x = "Hour", y = "Number of Tornadoes") +
  theme_minimal() 
```

##Casualties

```{r}
Torn <- All_Tornadoes %>%
  filter(yr >= 1994)
```
In 2018, there were 1126 tornadoes. 

```{r}
sum(Torn$fat, Torn$inj)
#27873 casualties
```

2018 tornadoes only: 
```{r}
states.sf <- st_transform(stateBorders, 
                            crs = st_crs(Torn))
gI <- st_intersects(Torn, states.sf, sparse = FALSE)
colnames(gI) <- states.sf$name
states.sf$GroupDayCount <- colSums(gI)
```

Remove Alaska, Hawaii, and Puerto Rico from the `states.sf` data frame.
```{r}
states.sf <- states.sf %>% 
  filter(!stusps %in% c("AK", "PR", "HI"))
```

Create a map of the outbreak density in each county in the United States. Fill by the `GroupDayCount` column. 
```{r}
states.sf <- st_transform(states.sf, 
                            crs = st_crs(stateBorders))
Torn <- st_transform(Torn, 
                            crs = st_crs(stateBorders))

```

```{r}
tm_shape(Torn) +
  tm_dots(alpha = 1, border.alpha = 1) +
tm_shape(states.sf, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_fill("GroupDayCount",
            title = "Tornado Counts",
            palette = "Reds", breaks = c(0, 500, 1000, 1500, 2000, 2500, 3000, 3500)) +
  tm_text("GroupDayCount", size = 0.75, just = c("left", "top")) +
  tm_borders(col = "gray", alpha = .3) +
  tm_compass(size = 5, position = c("right", "top")) + tm_scale_bar(width = 0.5, size = 1.4) +
  tm_layout(legend.position = c("right", "bottom"),
                   attr.position = c("left", "bottom"),
                   legend.frame = FALSE,
                   #title = "Tornado Group Days by County [1994-2017]",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
                   inner.margins = c(.1, .05, .05, 0.05),
            legend.width = -0.2, 
            legend.height = -0.3, 
            legend.title.size=1.25,
            legend.text.size = 1.1,
            legend.bg.color = "white",
            legend.bg.alpha =1) +
tm_shape(stateBorders, is.master = FALSE) +
  tm_borders() 
```


```{r}
totcas <- All_Tornadoes %>%
  group_by(st) %>%
  summarize(totaltorns = n(), 
    casualties = sum(cas)) 

#x <- sort(totcas, by = "st", decreasing=FALSE)

totcas <- st_transform(totcas, crs = st_crs(stateBorders))

totcas <- totcas %>%
  filter(st != "DC")
colnames(totcas) <- c("stusps", "totaltorns", "casualties", "geometry") 

test1 <- states.sf[order(states.sf$stusps),]
test2 <- totcas[order(totcas$stusps),]

test <- cbind(test1, totcas$casualties)



test <- st_sf(test, geom=test$geom)

states.sf <- st_as_sf(test, 
                      crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

st_geometry(states.sf) <- "geometry"
```

```{r}
tm_shape(totcas) +
  tm_dots(alpha = 0, border.alpha = 0) +
tm_shape(test, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_fill("totcas.casualties",
           title = "Tornado Counts",
           palette = "Reds", breaks = c(0, 1000, 2000, 3000, 4000, 5000)) +
 tm_text("totcas.casualties", size = 0.75, just = c("left", "top")) +
  tm_borders(col = "gray", alpha = .3) +
  tm_compass(size = 5, position = c("right", "top")) + tm_scale_bar(width = 0.5, size = 1.4) +
  tm_layout(legend.position = c("right", "bottom"),
                   attr.position = c("left", "bottom"),
                   legend.frame = FALSE,
                   #title = "Tornado Group Days by County [1994-2017]",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
                   inner.margins = c(.1, .05, .05, 0.05),
            legend.width = -0.2, 
            legend.height = -0.3, 
            legend.title.size=1.25,
            legend.text.size = 1.1,
            legend.bg.color = "white",
            legend.bg.alpha =1) +
tm_shape(stateBorders, is.master = FALSE) +
  tm_borders() 
```

```{r}
#save(All_Tornadoes, BigDays.sfdfT, BigDayTornadoes, file = "BigDays.RData")
load("BigDays.RData")
```

```{r}
BigDayTornadoes <- All_Tornadoes %>% 
  filter(groupNumber %in% BigDays.sfdfT$groupNumber,
         cDate %in% BigDays.sfdfT$cDate)
```

Figure of 1994-2000 and 2010-2018 county maximized outbreak activity

```{r}
counties.sf <- st_transform(counties.sf, 
                            crs = st_crs(BigDays.sfdfT))
gI <- st_intersects(BigDays.sfdfT, counties.sf, sparse = FALSE)
colnames(gI) <- counties.sf$name
counties.sf$GroupDayCount <- colSums(gI)
```

```{r}
BigDayTornadoes <- All_Tornadoes %>%
  filter(groupNumber %in% BigDays.sfdfT$groupNumber, 
         cDate %in% BigDays.sfdfT$cDate)
```

```{r}
tm_shape(BigDayTornadoes) +
  tm_symbols("mo")
```

Figure~\ref{May30} is an example of a big day in a large cluster. There were 88 tornadoes on that day. The cluster is identified as the eighth most prolific by our method (and the first most prolific by \cite{Forbes2006}) and extended over a two convective day period beginning on May 30th. This is the seventh largest of our big days as defined by the number of tornadoes in any large cluster.

Remove Alaska, Hawaii, and Puerto Rico from the `states.sf` data frame.
```{r}
states.sf <- states.sf %>% 
  filter(!stusps %in% c("AK", "PR", "HI"))
```

**Create a map showing the frequency of big tornado days by county. First transform the CRS of the county boundaries to match the CRS of `BigDays.sfdfT`.**
```{r}
counties.sf <- st_transform(counties.sf, 
                            crs = st_crs(BigDays.sfdfT))

stateBorders <- st_transform(stateBorders, 
                            crs = st_crs(BigDays.sfdfT))
```

**Extract the geographic centroids of tornado genesis on big tornado days by creating an `sf` object using `st_centroid()` that reduces the MULTIPOINT geometry to a POINT geometry. Compute the area of the group and the number of tornadoes per area (density).**
```{r}
groupDayCentroids.sfdfT <- st_centroid(BigDays.sfdfT)
groupDayCentroids.sfdfT$groupArea <- st_area(st_convex_hull(BigDays.sfdfT))
groupDayCentroids.sfdfT$groupDensity <- groupDayCentroids.sfdfT$nT/groupDayCentroids.sfdfT$groupArea
```

*Extract the Hull, centroids, and tornadoes for the May 30, 2004 big day. *

```{r}
May27 <- BigDays.sfdfT %>%
  filter(cDate == "1996-05-27")
May27 <- st_convex_hull(May27)

May27centroid <- groupDayCentroids.sfdfT %>%
  filter(cDate == "1996-05-27")

May27tornadoes <- BigDayTornadoes %>% 
  filter(groupNumber == May27centroid$groupNumber)

May27tornadoes <- May27tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour))
```

**Make a map of the May 30 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
May27tornadoes$Hour2 <- cut(May27tornadoes$Hour2, breaks=c(6,12,18,24,30))

May27 <- (tm_shape(stateBorders) + 
  tm_borders(col = "gray70", alpha = 1) +
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") + 
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("right", "top"), 
            inner.margins = c(.15, .15, .1, .2)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(May27, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(May27tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(May27centroid) +
  tm_symbols(size = 1.25, col = "black", shape = 24) )
```

```{r}
May05 <- BigDays.sfdfT %>%
  filter(cDate == "2007-05-05")
May05 <- st_convex_hull(May05)

May05centroid <- groupDayCentroids.sfdfT %>%
  filter(cDate == "2007-05-05")

May05tornadoes <- BigDayTornadoes %>% 
  filter(groupNumber == May05centroid$groupNumber)

May05tornadoes <- May05tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour))
```

**Make a map of the May 30 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
May05tornadoes$Hour2 <- cut(May05tornadoes$Hour2, breaks=c(6,12,18,24,30))

May05 <- (tm_shape(stateBorders) + 
  tm_borders(col = "gray70", alpha = 1) +
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") + 
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("right", "top"), 
            inner.margins = c(.15, .15, .1, .2)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(May05, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(May05tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(May05centroid) +
  tm_symbols(size = 1.25, col = "black", shape = 24) )
```


```{r}
tmap_arrange(May27, May05)
```

`Figure 2: Tornadoes occurring on May 30, 2004 as part of a big day within a large cluster. Each point represents a genesis location and is colored by the hour it occurred. The black triangle is the geographic center of the genesis locations. The black line is the minimum convex polygon around the genesis locations (convex hull).` #CORRECT

Create a map of the outbreak density in each county in the United States. Fill by the `GroupDayCount` column. 
```{r}
tm_shape(BigDays.sfdfT, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_polygons(alpha = 0, border.alpha = 0) +
tm_shape(counties.sf) +
   tm_fill("GroupDayCount",
            title = "Total Big Days",
            palette = "Reds", breaks = c(0, 20, 40, 60, 80, 100)) +
  tm_borders(col = "gray", alpha = .3) +
  tm_compass(size = 5, position = c("right", "top")) + tm_scale_bar(width = 0.6, size = 1.4) +
  tm_layout(legend.position = c("right", "bottom"),
                   attr.position = c("left", "bottom"),
                   legend.frame = FALSE,
                   #title = "Tornado Group Days by County [1994-2017]",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
                   inner.margins = c(.15, .0, .0, 0),
            legend.width = -0.2, 
            legend.height = -0.3, 
            legend.title.size=1.5,
            legend.text.size = 1.25,
            legend.bg.color = "white",
            legend.bg.alpha =1) +
tm_shape(stateBorders, is.master = FALSE) +
  tm_borders() 
```


\textit{FIGURE OF A CLUSTER}

Oceanic Nino Index: 
```{r}
#https://psl.noaa.gov/data/correlation/oni.data
```

Nino 3.4 
https://psl.noaa.gov/data/correlation/nina34.data

Average monthly teleconnection data
```{r}
#ftp://ftp.cpc.ncep.noaa.gov/wd52dg/data/indices/tele_index.nh
```

Solar Flux 
```{r}
#https://psl.noaa.gov/data/correlation/solar.data
```

Global Mean Land and Ocean Temperature Anomaly Index. More info here: https://data.giss.nasa.gov/gistemp/ 
```{r}
#https://psl.noaa.gov/data/correlation/gmsst.data
```

CMAP is the precip data sets. 
air.mon.anom.nc <- NOAA Global Surface Temperature

